---
title: "Bayesian Dose Response Modeling and Analysis of Four Kappa Opioid Receptor Antagonists"
author: "Madeline Martin"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dose_Response_Kappa_Opioid_Receptor_Antagonists}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, echo=FALSE}
# inspired by https://www.jumpingrivers.com/blog/knitr-default-options-settings-hooks/
knitr::opts_chunk$set(
  cache = TRUE,
  echo = FALSE,
  fig.path = "model_KOR_Antagonists_figures/knitr-",
  fig.retina = 2, # Control using dpi
  fig.width = 6,  # generated images
  fig.height = 5, # generated images
  fig.pos = "t",  # pdf mode
  fig.align = "center",
  dpi = if (knitr::is_latex_output()) 72 else 300, 
  out.width = "100%")
#  dev = "svg",
#  dev.args = list(png = list(type = "cairo-png")),
#  optipng = "-o1 -quiet")
# cmdstanr is more up-to-date than rstan and runs a little faster
# but can't expose defined functions for downstream analysis
if(Sys.info()["machine"] == "arm64"){
  #rstan does not currently support apple M1 chips
  stan_backend <- "cmdstanr"
} else {
  stan_backend <- "rstan"
}
```

```{r load-packages}
library(tidymodels)
library(BayesPharma)
```

---
# Dose Response Analysis of Four Novel Kappa Opioid Receptor Antagonists

In this case study we are going to reanalyze the dose response of 4 kappa opioid 
receptor (KOR) antagonists using the `BayesPharma` package from from a study 
done by Margolis et al. (2020). Whole cell electrophysiology in acute rat 
midbrain slices was used to evaluate pharmacological properties of four novel 
KOR antagonists: BTRX-335140, BTRX-395750, PF-04455242, and JNJ-67953964

In the paper, the dose-response data analysis was done by using the drc package
in R which implements the minimization of negative log likelihood function and 
reduces to least square estimation for a continuous response. The data was 
normalized to % baseline then fit to a 4-parameter log-logistic dose response 
model, setting the top (max response) to 100% and estimating the IC50, its 
variance, and the bottom (min response).

---
First, we will load the normalized dose-response data and plot it:

```{r}

load("../data/kor_antag.rda")

# The BayesPharma uses columns log_dose and response
kor_antag <- kor_antag %>%
  dplyr::mutate(
    response = normalized_measurement)
kor_antag
```

```{r plot-data}
ggplot2::ggplot(kor_antag) +
  ggplot2::theme_bw() +
  ggplot2::geom_point(
    mapping = ggplot2::aes(
      x = log_dose,
      y = response),
    size = 1.0,
    color = "black") +
  ggplot2::labs(title = "",
       x = "Dose(nM)",
       y = "% of baseline U-69,593 effect", 
       color = "Drug") + 
  ggplot2::facet_wrap(
    facets = dplyr::vars(Drug)) 
```


## Setting the Priors for the Model

For the priors, we are going to use a normal distribution because the response
values are continuous. 
First, We will run the analysis with top (max response) parameter prior set to 
a constant value of 100 because top is normalized to 100 and the default broad
priors for ic50, hill and bottom. Broad priors represent unbiased uncertainty 
and provide an opportunity for extreme responses.

Then, we will fit another model with more informative priors using the 
information from the research paper, and compare the two models. 

The level of informativeness of the priors will affect how much influence the
priors have on the model. Here is more [information on prior choice 
recommendations.](https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations)

```{r}
kor_priors <- BayesPharma::dr_priors(
  inhibitor = TRUE,
  top = 100)
kor_priors
```

### Initial Values for Parameters

Then, we will set inits to values close to parameter estimates to help with 
model convergence. The inits are the starting values for the mcmc chains of each 
parameter. The default inits: ec50 = -9, hill = -1, top = 100, bottom = 0

```{r, warning=FALSE, message=FALSE}
kor_inits <- BayesPharma::dr_inits()
```

### Prior Predictive Check

Before running the model, we will verify that the prior distributions
cover a plausible range of values for each parameter. To do this, we want to 
sample only from the prior distributions by adding  *sample_prior = "only"* as 
an argument to the `dr_model` function.
We will use the default response distribution of the model 
(*family = gaussian()*).

```{r, results='hide', message=FALSE}
kor_sample_priors <- BayesPharma::dr_model(
  kor_antag,
  BayesPharma::dr_formula(), 
  priors = kor_priors,
  init = kor_inits,
  sample_prior = "only")
```

Plot of the Prior Distributions:
```{r}
BayesPharma::density_distributions(
  kor_sample_priors, 
  half_max_label = "ic50",
  title_label = "Prior Density Distributions")
```


We can use this to evaluate the prior distributions for each parameter. The 
priors appear to cover a broad range of response values including extreme 
responses.


## Fitting the Model

To fit the model for each antagonist, we set the *predictors* argument in 
`dr_formula` to be *predictors = 0 + Drug*. 

```{r, results='hide', message=FALSE}
kor_model <- BayesPharma::dr_model(
  data = kor_antag,
  formula = BayesPharma::dr_formula(0 + Drug), 
  priors = kor_priors,
  inits = kor_inits)
```

### Results from the Model

```{r}
kor_model
```

## Checking for Convergence 
The model ran without warning messages meaning there were no parameter value 
problems or mcmc conflicts. The bulk and tail ESS indicate high resolution and 
stability. The R-hat for each parameter equals 1.00 and the traceplot shows the 
chains mixed well indicating the chains converged.

```{r}
BayesPharma::traceplot(
  kor_model, 
  predictors_col_name = "Drug", 
  half_max_label = "ic50") 
```

## Posterior Predictive Check

We plot a posterior predictive check to check if the data generated from the 
posterior distribution fits the observed data. The observed data is the dark 
blue line and the simulated data are the light blue lines. 

The simulated data fits some antagonist models better than others. The models 
for BTRX-395750 and JNJ fits the structure of the data decently well, the model 
for BTRX-335140 fits well but has a higher level of uncertainty, and the 
simulated data for the PF model is less structured, is an okay fit, and has
the highest level of uncertainty of the 4 drugs.

```{r}
BayesPharma::plot_pp_check(
  kor_model,
  plot_type = "dens_overlay_grouped",
  group = 'Drug',
  n = 50)

```

## Visualizing and Analyzing Results

Displayed below is a plot for the prior and posterior distributions of the 
parameters (prior is pink and posterior is teal). this can be useful for 
comparing the density distribution of the prior and posterior.
produced by the model:

```{r}
BayesPharma::prior_posterior_densities(
  kor_model,
  predictors_col_name = "Drug",
  half_max_label = "ic50")
```

Displayed below is a plot of the posterior distributions for each parameter 
with the confidence intervals and mean. This is a useful visual of the model 
results and to highlight the mode and high density intervals: 

```{r, message=FALSE, warning=FALSE}
BayesPharma::posterior_densities(
  kor_model, 
  predictors_col_name = "Drug", 
  half_max_label = "ic50")
```

Displayed below is a plot of a sample of 100 sigmoid dose-response curves from 
the posterior distribution (purple) and the median quantile intervals: 

```{r}
BayesPharma::posterior_draws_plot(
  model = kor_model,
  data = kor_antag, 
  predictors_col_name = "Drug",
  lower = -12,
  upper = -4,
  n = 50, 
  facet_var = Drug,
  title = "Dose-Response Posterior Draws",
  xlab = "Log[Molar]",
  ylab = "Response")
```


---
## Improving the Model with More Informative Priors

Below are the dose-response results discussed in the paper:
BTRX-335140: ic50 = 1.2 +/- 0.9 nM (-8.9 +/- 0.2 M), bottom = 1.3%
BTRX-395750: ic50 = 1.2 +/- 1.3 nM (-8.9 +/- 0.3 M), bottom = 28.0%
JNJ-67953964: ic50 = 3.0 +/- 4.6 nM (-8.5 +/- 0.4 M), bottom = -27%
PF-04455242: ic50 = 6.7 +/- 15.1 nM (-8.2 +/- 0.5 M), bottom = 45%

```{r}
kor_priors2 <- BayesPharma::dr_priors(
  ec50 = brms::prior(normal(-8.5, 0.5), nlpar = "ec50"),
  hill = brms::prior(normal(-1, 0.5), ub = 0.01, nlpar = "hill"),
  top = 100,
  bottom = brms::prior(normal(10, 15), nlpar = "bottom"))
kor_priors2
```

Sampling the Priors:

```{r, results='hide', message=FALSE}
kor_sample_priors <- BayesPharma::dr_model(
  kor_antag,
  BayesPharma::dr_formula(), 
  priors = kor_priors2,
  inits = kor_inits,
  sample_prior = "only")
```

Plot of the Prior Distributions:
```{r}
BayesPharma::density_distributions(
  kor_sample_priors,
  predictors_col_name = "_Intercept",
  half_max_label = "ic50")
```

Running the Model:

```{r, results='hide', message=FALSE}
kor_model2 <- dr_model(
  kor_antag,
  formula = BayesPharma::dr_formula(0 + Drug), 
  priors = kor_priors2,
  inits = kor_inits,
  cores = 2)

kor_model2
```


Checking the Model Convergence:

```{r}
BayesPharma::traceplot(kor_model2, "Drug", "ic50") 
```

Posterior Predictive Check:

```{r}
BayesPharma::plot_pp_check(
  kor_model2,
  "dens_overlay_grouped", 
  group = 'Drug',
  n = 50)
```

Plot of the Prior and Posterior Distributions:

```{r}
BayesPharma::prior_posterior_densities(
  kor_model2, 
  predictors_col_name = "Drug",
  half_max_label = "ic50")
```

Plot of the Posterior Distributions with the Mean and Confidence Intervals:

```{r, message=FALSE, warning=FALSE}
BayesPharma::posterior_densities(
  kor_model2, 
  predictors_col_name = "Drug",
  half_max_label = "ic50")
```

Sample of Sigmoid Dose-Response Curves from the Posterior Distribution and 
Mean Curve:

```{r}
BayesPharma::posterior_draws_plot(
  model = kor_model2,
  data = kor_antag, 
  predictors_col_name = "Drug",
  lower = -12,
  upper = -4,
  n = 50, 
  facet_var = Drug, 
  title = "Dose-Response Posterior Draws",
  xlab = "Log[Molar]",
  ylab = "Response")
```

## Comparing the Two Models Using LOO-Comparison:

The model with an elpd_diff and an se_diff of 0 is the preferred model. 
The kor_model2, the model with more informative priors, is the preferred model,
but only by a minute amount.

```{r}
kor_model <- BayesPharma::add_loo_criterion(kor_model)
kor_model2 <- BayesPharma::add_loo_criterion(kor_model2)

compare_models(kor_model, kor_model2)
```


---
##Analysis Using the drc Package

Here we will analyze the KOR antagonist data using the drc package and compare
it to the results from the BayesPharma analysis. 

We will fix the top to 100 and fit the ic50, hill, and bottom.

```{r load_packages}
library("drc") 
# https://rstats4ag.org/dose-response-curves.html#one-dose-response-curve
```

```{r}

BTRX_335140_df <- kor_antag[grepl("BTRX_335140", kor_antag$Drug),]
BTRX_395750_df <- kor_antag[grepl("BTRX_395750", kor_antag$Drug),]
JNJ_df <- kor_antag[grepl("JNJ", kor_antag$Drug),]
PF_df <- kor_antag[grepl("PF", kor_antag$Drug),]
```

Here is the code and results of the drc dose response model:
```{r}
BTRX_335140_drc <- drc::drm(
  response ~ log_dose,
  data = BTRX_335140_df,
  fct = L.4(fixed = c(NA, NA, 100, NA),
  names = c("hill", "bottom", "top", "ic50")))

BTRX_395750_drc <- drc::drm(
  response ~ log_dose,
  data = BTRX_395750_df,
  fct = L.4(fixed = c(NA, NA, 100, NA),
  names = c("hill", "bottom", "top", "ic50")))

JNJ_drc <- drc::drm(
  response ~ log_dose,
  data = JNJ_df,
  fct = L.4(fixed = c(NA, NA, 100, NA),
  names = c("hill", "bottom", "top", "ic50")))

PF_drc <- drm(
  response ~ log_dose,
  data = PF_df,
  fct = L.4(fixed = c(NA, NA, 100, NA),
  names = c("hill", "bottom", "top", "ic50")))

summary(BTRX_335140_drc)
summary(BTRX_395750_drc)
summary(JNJ_drc)
summary(PF_drc)

```

Displayed below is a plot of the results for KOR antagonist BTRX_335140 with
the data points, mean and error bars of the data points at each concentration 
they were recorded at, and a fitted curve.
```{r, echo=FALSE}

plot(BTRX_335140_drc,
     log = "", 
     type = c("average"), 
     xlim = c(-11, -6),
     ylim = c(-50,150),
     col = "red",
     pch = 16)
plot(BTRX_335140_drc, 
     add = TRUE,
     log = "", 
     type = c("all"), 
     xlim = c(-11, -6),
     ylim = c(-50,150),
     col = "gray",
     pch = 16)
plot(BTRX_335140_drc, 
     add = TRUE,
     log = "", 
     type = c("bars"), 
     xlim = c(-11, -6),
     ylim = c(-50,150),
     col = "black",
     lty = 1)
```

Displayed below are plots of the curves from drc and BayesPharma for each KOR 
antagonists. Here we can see how the fits from the different methods 
differentiate. 


```{r}
log_dose <- seq(min(-4), max(-12), length.out=100)

BTRX_335140_pred_value <- predict(BTRX_335140_drc, expand.grid(log_dose, 1))
BTRX_335140_drc_df <- data.frame(log_dose, BTRX_335140_pred_value)

BTRX_395750_pred_value <- predict(BTRX_395750_drc, expand.grid(log_dose, 1))
BTRX_395750_drc_df <- data.frame(log_dose, BTRX_395750_pred_value)

JNJ_pred_value <- predict(JNJ_drc, expand.grid(log_dose, 1))
JNJ_drc_df <- data.frame(log_dose, JNJ_pred_value)

PF_pred_value <- predict(PF_drc, expand.grid(log_dose, 1))
PF_drc_df <- data.frame(log_dose, PF_pred_value)


```
```{r, echo=FALSE}
plot_drc <- function(drc_df, 
                     pred_value,
                     data,
                     title,
                     values = colors) {
  ggplot2::ggplot(drc_df) +
    ggplot2::geom_line(mapping = ggplot2::aes(x = log_dose, 
                                              y = pred_value,
                                              color = "drc"),
                       size = 0.6) +
    ggplot2::geom_point(data = data, 
                        mapping = ggplot2::aes(x = log_dose, 
                                               y = response)) +

    ggplot2::labs(title = title,
                  color = "Fit") + 
    ggplot2::xlab("log dose") +
    ggplot2::ylab("response") +
    ggplot2::scale_color_manual(values = values)
}

kor_post_means <- BayesPharma::posterior_mean(kor_model2,
                                              n = 100,
                                              lower = -12,
                                              upper = -4,
                                              predictor_name = "Drug") 

colors = c("drc" = "blue", "Bayes" = "red")

BTRX_335140_drc_plot <- plot_drc(BTRX_335140_drc_df, 
                                 BTRX_335140_pred_value,
                                 BTRX_335140_df,
                                 "BTRX_335140") +
  ggplot2::geom_line(data = kor_post_means[grepl("BTRX_335", 
                                                 kor_post_means$Drug),],
                     mapping = ggplot2::aes(x = log_dose, 
                                            y = Response,
                                            color = "Bayes"),
                     size = 0.6) +
  ggplot2::ylab("% Baseline") +
  ggplot2::xlab("Log[Molar] EM1")

BTRX_395750_drc_plot <- plot_drc(BTRX_395750_drc_df, 
                                 BTRX_395750_pred_value,
                                 BTRX_395750_df,
                                 "BTRX_395750") +
  ggplot2::geom_line(data = kor_post_means[grepl("BTRX_395", 
                                                 kor_post_means$Drug),],
                     mapping = ggplot2::aes(x = log_dose, 
                                            y = Response,
                                            color = "Bayes"),
                     size = 0.6) +
  ggplot2::ylab("% Baseline") +
  ggplot2::xlab("Log[Molar] EM1")

JNJ_drc_plot <- plot_drc(JNJ_drc_df, 
                         JNJ_pred_value,
                         JNJ_df,
                         "JNJ") +
  ggplot2::geom_line(data = kor_post_means[grepl("JNJ", 
                                                 kor_post_means$Drug),],
                     mapping = ggplot2::aes(x = log_dose, 
                                            y = Response,
                                            color = "Bayes"),
                     size = 0.6) +
  ggplot2::ylab("% Baseline") +
  ggplot2::xlab("Log[Molar] EM1")

PF_drc_plot <- plot_drc(PF_drc_df, 
                        PF_pred_value,
                        PF_df,
                        "PF") +
  ggplot2::geom_line(data = kor_post_means[grepl("PF", 
                                                 kor_post_means$Drug),],
                     mapping = ggplot2::aes(x = log_dose, 
                                            y = Response,
                                            color = "Bayes"),
                     size = 0.6) +
  ggplot2::ylab("% Baseline") +
  ggplot2::xlab("Log[Molar] EM1")

bayesplot::bayesplot_grid(plots = list(BTRX_335140_drc_plot,
                                       BTRX_395750_drc_plot,
                                       JNJ_drc_plot,
                                       PF_drc_plot)) 
  
```

```{r, echo=FALSE}

drc_stats <- function(drc_fit, drug) {
  
  data.frame(variables = c("hill", "bottom", "ic50")) %>%
    dplyr::mutate(
      model = "drc",
      drug_label = drug,
      mean = summary(drc_fit)$coefficients[, 1],
      std_error = summary(drc_fit)$coefficients[, 2],
      l_ci = summary(drc_fit)$coefficients[, 1] - 
        summary(drc_fit)$coefficients[, 2],
      u_ci = summary(drc_fit)$coefficients[, 1] + 
        summary(drc_fit)$coefficients[, 2],
      t_value = summary(drc_fit)$coefficients[, 3],
      p_value = summary(drc_fit)$coefficients[, 4]
      )
  
}

BTRX_335140_drc_stats <- drc_stats(BTRX_335140_drc, "BTRX_335140")

BTRX_395750_drc_stats <- drc_stats(BTRX_395750_drc, "BTRX_395750")

JNJ_drc_stats <- drc_stats(JNJ_drc, "JNJ")

PF_drc_stats <- drc_stats(PF_drc, "PF")

```

```{r, echo=FALSE}
BTRX_335140_drc_stats[3,][BTRX_335140_drc_stats[3,]$variables == "ic50", "variables"] <- "ic50_BTRX_335140"
BTRX_395750_drc_stats[3,][BTRX_395750_drc_stats[3,]$variables == "ic50", "variables"] <- "ic50_BTRX_395750"
JNJ_drc_stats[3,][JNJ_drc_stats[3,]$variables == "ic50", "variables"] <- "ic50_JNJ"
PF_drc_stats[3,][PF_drc_stats[3,]$variables == "ic50", "variables"] <- "ic50_PF"

drc_ic50 <- dplyr::bind_rows(BTRX_335140_drc_stats[3,], 
          BTRX_395750_drc_stats[3,],
          JNJ_drc_stats[3,],
          PF_drc_stats[3,])

BTRX_335140_drc_stats[1,][BTRX_335140_drc_stats[1,]$variables == "hill", "variables"] <- "hill_BTRX_335140"
BTRX_395750_drc_stats[1,][BTRX_395750_drc_stats[1,]$variables == "hill", "variables"] <- "hill_BTRX_395750"
JNJ_drc_stats[1,][JNJ_drc_stats[1,]$variables == "hill", "variables"] <- "hill_JNJ"
PF_drc_stats[1,][PF_drc_stats[1,]$variables == "hill", "variables"] <- "hill_PF"

drc_hill <- dplyr::bind_rows(BTRX_335140_drc_stats[1,], 
                             BTRX_395750_drc_stats[1,],
                             JNJ_drc_stats[1,],
                             PF_drc_stats[1,])

BTRX_335140_drc_stats[2,][BTRX_335140_drc_stats[2,]$variables == "bottom", "variables"] <- "bottom_BTRX_335140"
BTRX_395750_drc_stats[2,][BTRX_395750_drc_stats[2,]$variables == "bottom", "variables"] <- "bottom_BTRX_395750"
JNJ_drc_stats[2,][JNJ_drc_stats[2,]$variables == "bottom", "variables"] <- "bottom_JNJ"
PF_drc_stats[2,][PF_drc_stats[2,]$variables == "bottom", "variables"] <- "bottom_PF"

drc_bottom <- dplyr::bind_rows(BTRX_335140_drc_stats[2,], 
                               BTRX_395750_drc_stats[2,],
                               JNJ_drc_stats[2,],
                               PF_drc_stats[2,])

```

Displayed below is the comparison of results from drc and BayesPharma for each 
parameter of the dose-response curve. Here we see that the Bayesian method 
provides a distribution curve as evidence and has smaller confidence intervals 
than most of the standard errors provided by the drc method. 

```{r, echo=FALSE}
bayes_vs_drc <- function(model,
                         filter_label, 
                         predictors_col_name = "_Intercept",
                         half_max_label = NULL,
                         drc) {
  posterior <- dplyr::bind_rows(
    model %>%
      tidybayes::tidy_draws() %>%
      tidybayes::gather_variables() %>%
      dplyr::filter(!stringr::str_detect(.variable, "__$")) %>%
      dplyr::filter(!stringr::str_detect(.variable, "sigma")) %>%
      dplyr::filter(stringr::str_detect(.variable, filter_label))
  ) %>%
    dplyr::rename(variables = .variable) %>%
    dplyr::mutate(variables = stringr::str_extract(variables,
                                                   "b_[a-zA-Z0-9]+.{1,100}") %>%
                    stringr::str_remove("b_")) %>%
    dplyr::mutate(variables = stringr::str_extract(variables,
                                                   "[a-zA-Z0-9]+.{1,100}") %>%
                    stringr::str_remove(predictors_col_name)) %>%
    dplyr::mutate(variables = stringr::str_extract(variables,
                                                   "[a-zA-Z0-9]+.{1,100}") %>%
                    stringr::str_replace(filter_label, half_max_label))
  
  model <- c("Bayes" = "blue", "drc" = "red")

  ggplot2::ggplot() +
    ggplot2::theme_bw() +
    ggplot2::theme(legend.position = "right") +
    tidybayes::stat_halfeye(data = posterior,
                  ggplot2::aes(x = .value, y = variables, color = "Bayes"), 
                  .width = c(0.66, 0.95)) +
    tidybayes::geom_pointinterval(data = drc,
                    ggplot2::aes(y = variables, x = mean, xmin = l_ci, 
                                 xmax = u_ci, color = "drc"),
                    position = ggplot2::position_nudge(y = -0.1),
                    size = 3) +
    ggplot2::labs(x = "Log[Molar] EM1", color = "model") +
    ggplot2::scale_color_manual(values = model)
}
```
```{r}

bayes_vs_drc(kor_model2, "ec50", "Drug", "ic50", drc_ic50) 

```

(The drc results for the hill are displayed as a positive value although the 
curve has a negative slope.)
```{r}

hill_bd1 <- bayes_vs_drc(kor_model2, "hill", "Drug", "hill", drc_hill) 
hill_bd2 <- bayes_vs_drc(kor_model2, "hill", "Drug", "hill", drc_hill) +
  ggplot2::xlim(-6, 14)

bayesplot::bayesplot_grid(hill_bd1, hill_bd2)
```

```{r}

bayes_vs_drc(kor_model2, "bottom", "Drug", "bottom", drc_bottom) 
```


