---
title: "Bayesian Dose Response Modeling and Analysis of Damgo"
author: "Madeline Martin"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dose_Response_Damgo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# cmdstanr is more up-to-date than rstan and runs a little faster
# but can't expose defined functions for downstream analysis
if(Sys.info()["machine"] == "arm64"){
  cat("rstan does not currently support apple M1 chips, setting stan_backend = 'cmdstanr'")
  stan_backend <- "cmdstanr"
} else {
  stan_backend <- "rstan"
}

```

```{r setup}
# library(BayesPharma)
```

---

# Case Study: Analyzing Dose Response of Damgo 

give some background on the data


```{r}

xl_damgo_dr <- "C:/Users/marti/Downloads/damgo_dr_reformatted.xlsx"

damgo <- readxl::read_excel(xl_damgo_dr)

```
```{r}
ggplot2::ggplot(damgo) +
  ggplot2::theme_bw() +
  ggplot2::geom_point(
    ggplot2::aes(x = dose_nM,
        y = normalized_response),
    size = 1.0,
    color = "black") +
  ggplot2::labs(title = "Damgo",
       x = "Dose (nM)",
       y = "% of baseline U-69,593 effect") + 
  ggplot2::scale_x_log10()
```

There are limitations to the package that requires specific column names and the
dose to be in log scale; therefore, it is necessary to use the 
`calculate_log_dose` and `change_col_names` functions before running the model. 

- Maybe unneeded V - 

The `calculate_log_dose` function requires the data as an input, the dose column
name, and the molar concentration. The function creates a new column `log_dose`.

The `change_col_names` function requires data, the response column name, the
log dose column name, and predictors column name (the column containing the
different perturbations tested). The `calculate_log_dose` function automatically
changes the column name to log_dose and for this model, there is only 1
perturbation being tested; therefore, the *log_dose_col_name* and
*predictors_col_name* can be left alone because the defaults for the column name
arguments are set to *NULL*.

- Maybe unneeded ^ -  

```{r}
damgo <- BayesPharma::calculate_log_dose(data = damgo, dose_col = dose_nM, 
                                         molar_concentration = 10^(-9))
damgo <- BayesPharma::change_col_names(data = damgo, 
                                       response_col_name = normalized_response)
damgo
```

## Setting the Priors for the Model

talk about priors for this model (find actual info about Damgo)
A previous study done on Damgo found an IC50 value of X, a minimal response 
(bottom) of X, and a hill slope of X.

max response (top) is normalized to 100, therefore, top parameter prior set to a 
constant value of 100. First, we will run the model with the default weakly 
informative priors, then build more informative priors based on the information 
provided by a previous study done on Damgo. 

The default weakly informative priors being used for ec50, hill, and bottom:
ec50: normal distribution with a mean of -7 and sd of 2.5
hill: normal distribution mean = -1, sd = 1, upper bound = 0.01
bottom: normal distribution mean = 0, sd = 25

```{r}
damgo_priors <- BayesPharma::dr_priors(inhibitor = TRUE, 
                                       top = 100)
damgo_priors
```

### Initial Values for Parameters
The inits are the starting values for the mcmc chains of each parameter. By 
setting inits to values close to parameter estimates can help with model 
convergence. 

The default inits for each parameter are: 
ec50 = -9
hill = -1 if inhibitor = TRUE or hill = 1 for inhibitor = FALSE
top = 100
bottom = 0

```{r}
damgo_inits <- BayesPharma::dr_inits(inhibitor = TRUE)
damgo_inits
```

### Sampling the Priors to Check if the Prior Distributions Cover Plausible 
### Parameter Values

```{r}
sampling_priors <- BayesPharma::dr_model(data = damgo,
                                         model_name = damgo1, 
                                         formula = dr_formula(),
                                         priors = damgo_priors,
                                         inits = damgo_inits,
                                         sample_prior = "only")

```
```{r}
BayesPharma::density_distributions(sampling_priors, half_max_label = "ic50")
```

The prior distributions cover the pluasible range of values in the form of a 
normal distribution. Next step is to run the model!


## Running the Bayesian Model
dr_model uses brms and rstan or cmdstanr

```{r}

damgo_model <- BayesPharma::dr_model(data = damgo,
                                     model_name = damgo1,
                                     formula = BayesPharma::dr_formula(),
                                     priors = damgo_priors,
                                     inits = damgo_inits)
```

### Results from the Model

```{r}
damgo_model
```

## Checking for Convergence 
The model ran without any warning messages, Rhat for all parameters is 1.00 
(NA for top because it was set to a constant value of 100), Bulk and Tail ESS 
are high, and the mcmc chains appear to have mixed well throughout the model 
run.

```{r}
damgo_trace <- BayesPharma::traceplot(damgo_model, 
                                      half_max_label = "ic50")
damgo_trace
```

```{r}
damgo_trace$data
```

## Posterior Predictive Check
This is used to check if the data generated from the posterior distribution 
fits the observed data. The observed data is the dark blue line and the 
simulated data are the light blue lines. 
The simulated data somewhat fits the observed data and contains outliers. 
This may be due to the weakly informative priors. 

```{r}
BayesPharma::plot_pp_check(damgo_model)
```

## Visualizing and Analyzing Results

Displayed below is a plot for the prior and posterior distributions of the 
parameters. Prior is pink and posterior is teal: 

```{r}
BayesPharma::prior_posterior_densities(damgo_model, half_max_label = "ic50")
```

Displayed below is a plot of the posterior distributions for each parameter 
with the confidence intervals and mean: 

```{r}
BayesPharma::posterior_densities(damgo_model, half_max_label = "ic50")
```

Displayed below is a plot of a sample of 50 sigmoid dose-response curves from 
the posterior distribution (purple) and the median quantile intervals: 


```{r}
posterior_draws_plot(model = damgo_model, data = damgo, predictors_col_name = "Damgo",
                     lower = -10, upper = -3, n = 50, facet_var = "Damgo", 
                     title = "Dose-Response Posterior Draws",
                     xlab = "Log[Molar]", ylab = "Response")

```

---

## Improving the model with more informative priors:

```{r}
damgo_prior2 <- BayesPharma::dr_priors(ec50 = brms::prior(normal(-6, 1.5), 
                                                          nlpar = "ec50"),
                                       hill = brms::prior(normal(-0.5, 0.5),
                                                          nlpar = "hill"),
                                                          top = 100,
                                       bottom = brms::prior(normal(12.5, 6.25), 
                                                            nlpar = "bottom")
                                       )
damgo_prior2
```

Sampling the Priors:

```{r}
sample_priors2 <- BayesPharma::dr_model(damgo, 
                                        damgo2,
                                        dr_formula(),
                                        priors = damgo_prior2,
                                        inits = damgo_inits,
                                        sample_prior = "only")

```


Plot of the Prior Distributions:

```{r}
BayesPharma::density_distributions(sample_priors2, half_max_label = "ic50")
```

Running the Model:

```{r}
damgo_model2 <- BayesPharma::dr_model(damgo, 
                                      damgo2,
                                      BayesPharma::dr_formula(),
                                      priors = damgo_prior2,
                                      inits = damgo_inits)
damgo_model2
```

Checking the Model Convergence:

```{r}
BayesPharma::traceplot(damgo_model2)
```

Posterior Predictive Check:

```{r}
BayesPharma::plot_pp_check(damgo_model2)
```

Plot of the Prior and Posterior Distributions:

```{r}
BayesPharma::prior_posterior_densities(damgo_model2)
```

Plot of the Posterior Distributions with the Mean and Confidence Intervals:

```{r}
BayesPharma::posterior_densities(damgo_model2)
```

Sample of Sigmoid Dose-Response Curves from the Posterior Distribution and 
Mean Curve:

```{r}
BayesPharma::plot_draws_data(model = damgo_model2,
                             n = 100,
                             lower = -9,
                             upper = -3,
                             predictor_name = "Damgo",
                             data = damgo,
                             measurement = damgo$response,
                             title = NULL,
                             xlabel = "Log[Molar]",
                             ylabel = "% Baseline U-69,593 Effect")
```

## Comparing the Two Models Using LOO-Comparison:

The model with the greater value is the preferred model. Model damgo2, the model 
with more informative priors, is the preferred model.

```{r}
damgo_model <- BayesPharma::add_loo_criterion(damgo_model, damgo1)
damgo_model2 <- BayesPharma::add_loo_criterion(damgo_model2, damgo2)
BayesPharma::compare_models(damgo_model, damgo_model2)
```

