---
title: "Modeling the Pnear Distribution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Modeling the Pnear Distribution}
  %\VignetteEncoding{UTF-8}
---

```{r set-options, echo=FALSE}

# inspired by https://www.jumpingrivers.com/blog/knitr-default-options-settings-hooks/
knitr::opts_chunk$set(
  cache = TRUE,
  echo = FALSE,
  fig.path = "model_Pnear_distribution_figures/knitr-",
  fig.retina = 2, # Control using dpi
  fig.width = 6,  # generated images
  fig.height = 5, # generated images
  fig.pos = "t",  # pdf mode
  fig.align = "center",
  dpi = if (knitr::is_latex_output()) 72 else 300, 
  out.width = "100%",
  dev = "svg",
  dev.args = list(png = list(type = "cairo-png")),
  optipng = "-o1 -quiet")

# cmdstanr is more up-to-date than rstan and runs a little faster
# but can't expose defined functions for downstream analysis
if(Sys.info()["machine"] == "arm64"){
  cat("rstan does not currently support apple M1 chips, setting stan_backend = 'cmdstanr'")
  stan_backend <- "cmdstanr"
} else {
  stan_backend <- "rstan"
}

```


```{r load-packages}
library(tidyverse)
library(BayesPharma)
library(xkcd)
library(GeomIndicator)
```
# Modeling Folding Funnels
A common task in molecular modeling is to predict the conformation of the folded
state for a given molecular system. For example, the Rosetta ab initio, or
protein-protein-interface docking protocols. To turn the simulation into a prediction
requires predicting the relative free energy of the folded state relative a reference.

The Rosetta score function can score individual conformations, but doesn't capture
the free energy of the state. Typically, a researcher will run a series of trajectories
and generate a **score vs. RMSD plot** and look for a "folding funnel" e.g. lower
energies for conformations that are closer to a target folded state. Here, RMSD
is the root-mean squared deviation measuring the euclidean distance of pairs of atom
defined by the application (for example just the backbone for sequence design or
interface atoms for docking).

## Pnear score
To quantify the quality of the folding funnel, recently, there has been interest
in using the **Pnear score**, which is defined by

    Pnear = Sum_i[exp(-RMSD[i]^2/lambda^2)*exp(-score[i]/k_BT)] /
            Sum_i[exp(-score[i]/k_BT)]

where *(RMSD[i], score[i])* is the score RMSD and score values for a
conformation *i*. The parameter *lambda* is measured in Angstroms indicating the
breadth of the Gaussian used to define "native-like-ness".  The bigger the
value, the more permissive the calculation is to structures that deviate from
native. Typical values for peptides range from 1.5 to 2.0, and for proteins from
2.0 to perhaps 4.0. And finally the parameter *k_BT* is measured in in energy
units, determines how large an energy gap must be in order for a sequence to be
said to favour the native state. The default value, 0.62, should correspond to
physiological temperature for ref2015 or any other scorefunction with units of
kcal/mol.

## Two state model
Thinking of the folded and unfolded states as a two-state model and RSMD as
a reaction coordinate or "collective variable", then the energy gap can be
modeled by a sigmoidal Boltzmann distribution.

```{r simulate-data}

simulate_score_vs_rmsd <- function(
  rmsd_min = 0,
  rmsd_max = 5,
  energy_near = 0,
  energy_far = 10,
  radius = .5,
  hill = 1,
  sigma = 5,
  n_samples = 300) {
  
  tibble::tibble(
    rmsd = runif(n_samples, min=rmsd_min, max=rmsd_max),
    log_rmsd = log10(rmsd),
    score_mean = 
      energy_near + 
      (energy_far - energy_near) / (1 + 10^((log(radius) - log_rmsd)*hill)),
    score = rnorm(
      n = n_samples,
      mean = score_mean,
      sd = sigma))
  }

data_simulated <- simulate_score_vs_rmsd(
  radius = 1.5,
  hill=1.8,
  energy_near = -30,
  energy_far = -10,
  sigma = 5) %>%
  dplyr::mutate(
    Pnear = BayesPharma::Pnear(
      score = score,
      rmsd = rmsd))

```

```{r sigmoidal-cartoon, dependson=c("load-packages", "simulate-data")}
ggplot2::ggplot(
  data = data_simulated) +
  xkcd::theme_xkcd() +
    ggplot2::ggtitle(
    "Score vs RMSD") +
  xkcd::xkcdaxis(
    xrange = c(0, 5),
    yrange = c(-35, -5)) +
  ggplot2::ylab("Score") +
  ggplot2::xlab(
    expression("RMSD "*ring(A)*"")) +
  ggplot2::geom_line(
    mapping = ggplot2::aes(
      x = rmsd,
      y = score_mean),
    size = 2,
    color = "blue")
```

For a principled molecular dynamics or monte carlo simulation that maintains detailed balance, it is in
theory possible to use thermodynamic integration to quantify the energy gap
between the two states. However, this is often not computationally feasible for
proteins of moderate size or in a protein design or screening context where many
different molecules need to be evaluated given a limited computational budget. So,
Instead, we will assume that the at least locally around the folded state, the
degrees of freedom increase exponentially so that the log of the RMSD defines
a linear reaction coordinate. 

If we simulate, trajectory points from the sigmoid on the log(RMSD) scale,
with a Normally distributed error we can generate synthetic score-vs-rmsd plots

```{r simulate-score-vs-rmsd-data, dependson=c("load-packages", "simulate-data")}

pnear_score <- BayesPharma::Pnear(
  score = data_simulated$score,
  rmsd = data_simulated$rmsd)

ggplot2::ggplot(data = data_simulated) + 
  ggplot2::theme_bw() +
  GeomIndicator::geom_indicator(
    mapping = ggplot2::aes(indicator = paste0("Pnear: ", signif(pnear_score, 2))),
    xpos = 0.05,
    ypos = 0.9,
    group = 0) +
  ggplot2::geom_line(
    mapping = ggplot2::aes(
      x = rmsd,
      y = score_mean),
    color = "blue") +
  ggplot2::geom_point(
    mapping = ggplot2::aes(
      x = rmsd,
      y = score)) +
  ggplot2::ggtitle(
    "Simulated score vs RMSD") +
  ggplot2::scale_y_continuous("Score (REU)") +
  ggplot2::scale_x_continuous(
    expression("RMSD ("*ring(A)*")"))
```

A nice thing about having the parametric model to generate score-vs-rmsd plots,
is that it allows us to measure measure the sensitivity of the Pnear to differently
shaped score-vs-rmsd plots. For example we can scann both the radius of 

```{r simulate-data-scan, dependson=c("load-packages", "simulate-data")}
data_simulated_scanned <- tidyr::expand_grid(
  radius = c(0.5, 1.5, 3),
  energy_near = c(-50, -30, -20)) %>%
  dplyr::rowwise() %>%
  dplyr::do({
    params <- .
   simulate_score_vs_rmsd(
      radius = params$radius,
      hill=1.8,
      energy_near = params$energy_near,
      energy_far = -10,
      sigma = 5) %>%
      dplyr::mutate(
        radius = params$radius,
        radius_label = paste0("Radius: ", params$radius),
        energy_near = params$energy_near,
        energy_near_label = paste0("Energy Near: ",params$energy_near),
        Pnear = BayesPharma::Pnear(
          score = score,
          rmsd = rmsd,
          kbt = 2))
  })

ggplot2::ggplot(data = data_simulated_scanned) + 
  ggplot2::theme_bw() +
  GeomIndicator::geom_indicator(
    mapping = ggplot2::aes(
      indicator = paste0("Pnear: ", signif(Pnear, 2))),
    xpos = 0.08,
    ypos = 0.8,
    group = 0,
    color = "darkgreen") +
  ggplot2::geom_line(
    mapping = ggplot2::aes(
      x = rmsd,
      y = score_mean),
    color = "blue") +
  ggplot2::geom_point(
    mapping = ggplot2::aes(
      x = rmsd,
      y = score),
    size = .5,
    alpha = .8) +
  ggplot2::facet_grid(
    cols = dplyr::vars(radius_label),
    rows = dplyr::vars(energy_near_label)) +
  ggplot2::ggtitle(
    "Simulated score vs RMSD") +
  ggplot2::scale_y_continuous("Score (REU)") +
  ggplot2::scale_x_continuous(
    expression("Interface RMSD ("*ring(A)*")"))

```

Another question we can use this model to investigate is how reproducible is the Pnear score?

```{r simulated-replicates, dependson=c("load-packages", "simulate-data")}
data_simulated_replicates <- tidyr::expand_grid(
  radius = c(1.5),
  energy_near = c(-30),
  lambda = c(0.5, 1.5, 2),
  kbt = c(0.62, 2, 5),
  replicate = seq(1, 5000)) %>%
  dplyr::rowwise() %>%
  dplyr::do({
    params <- .
    conformations <- simulate_score_vs_rmsd(
      radius = params$radius,
      hill=1.8,
      energy_near = params$energy_near,
      energy_far = -10,
      sigma = 5)
    data.frame(
      replicate = params$replicate,
      Pnear = BayesPharma::Pnear(
        score = conformations$score,
        rmsd = conformations$rmsd,
        lambda = params$lambda,
        kbt = params$kbt),
      lambda = params$lambda,
      lambda_label = paste0("Lambda: ", params$lambda),
      kbt_label = paste0("k_BT: ", params$kbt))
  })

params <- data.frame(
  parameter = c(
    "Energy Near",
    "Energy Far",
    "Radius",
    "Hill"),
  value = c(-30, -10, 1.5, 1.8))
  
ggplot2::ggplot(data = data_simulated_replicates) +
  ggplot2::theme_bw() +
  ggplot2::ggtitle(
    label = "Pnear scores for Simulated Score vs. RMSD") +
  ggplot2::facet_grid(
    cols = dplyr::vars(lambda_label),
    rows = dplyr::vars(kbt_label)) +
  # GeomIndicator::geom_indicator(
  #   data = params,
  #   mapping = ggplot2::aes(
  #     indicator = paste0(parameter, ": ", value),
  #     group = parameter),
  #   xpos = .02,
  #   ypos = "top") +
  ggplot2::geom_histogram(
    mapping = ggplot2::aes(
      x = Pnear),
      fill = "darkgreen",
      bins = 150)
```  

# Antibody SnugDock Case study
As a case study, we can look at the real score-vs-rmsd plots from the 
Antibody SnugDock scientific benchmark. It is evaluates the SnugDock protocol over 6
Antibody protein targets

```{r snugdock-load-data, dependson=c("load-packages"), warning=FALSE, message=FALSE}

data <- data.frame(
  target = c("1ahw", "1jps", "1mlc", "1ztx", "2aep", "2jel")) %>%
  dplyr::rowwise() %>%
  dplyr::do({
    target <- .$target
    readr::read_table(
      file = system.file(
        "extdata", "Pnear", "antibody_snugdock", paste0(target, "_rmsd.sc"),
        package = "BayesPharma",
        mustWork = TRUE),
      skip = 1) %>%
      dplyr::mutate(target = target, .before = 1) %>%
      dplyr::select(target, rmsd = Irms, score = I_sc)
  }) %>%
  dplyr::filter(rmsd <= 20) %>%
  dplyr::mutate(
    response = score,
    log_dose = log(rmsd))

pnear_scores <- data %>%
  dplyr::group_by(target) %>%
  dplyr::do({
    data_by_target <- .
    data.frame(
      Pnear = BayesPharma::Pnear(
        score = data_by_target$score,
        rmsd = data_by_target$rmsd))
  }) %>%
  dplyr::mutate(
    Pnear_label = Pnear %>% signif(2))

```

We can use the fit the sigmoid model to the log(RMSD) using the
BayesPharma package, which relies on BRMS and Stan

```{r stugdock-fit-model, output=FALSE, dependson=c("load-packages", "stugdock-load-data"), message=FALSE, warning=FALSE, refresh = 0, results="hide"}
model <- data %>%
  BayesPharma::dr_model(
    formula = BayesPharma::dr_formula(
      multiple_perturbations = TRUE,
      predictors = 0 + target),
    prior = BayesPharma::dr_priors(
      ec50 = brms::prior(normal(1.5,2), nlpar = "ec50"),
      top = brms::prior(normal(-10,10), nlpar = "top"),
      bottom = brms::prior(normal(-20,10), nlpar = "bottom"),
      inhibitor = FALSE),
    chains = 4, cores = 10, backend = stan_backend,
    silent = 1)

if(stan_backend == "cmdstanr") {
  sigmoid <- Vectorize(function(ec50, hill, top, bottom, log_dose){
    bottom + (top - bottom) / (1 + 10^((ec50 - log_dose)*hill))
  })
} else if (stan_backend == "rstan") {
  model %>% brms::expose_functions()
} else {
  stop("Unrecognized stan backend '", stan_backend, "\n", sep = "")
}

```

Check the model parameter fit and estimated parameters:
```{r mode-fit-summary, dependson=c("load-packages", "stugdock-fit-model")}
model
```

Excitingly, using leave-one-out cross-validation, the sigmoid model fits the
data very well
```{r model-fit-loo, dependson=c("load-packages", "stugdock-fit-model")}
model <- model %>% brms::add_criterion(criterion = "loo")
model$criteria$loo
```



```{r snugdock-collect-draws,  dependson=c("load-packages", "stugdock-fit-model")}
# collect posterior draws


lower <- .01
upper <- 3
n <- 100

# Samples of the expected response across RMSD values for draws from the posterior
ep_data <- model %>%
  tidybayes::add_epred_draws(
    newdata = tidyr::expand_grid(
      log_dose = seq(from = lower, to = upper, length.out = 100),
      target = data$target %>% unique()),
    value = "response",
    ndraws = 30) %>%
  dplyr::mutate(
    rmsd = exp(log_dose),
    score = response)

# Median quantile intervals for the response across RMSD values for draws from the posterior
pp_data <- model %>%
  tidybayes::add_predicted_draws(
    newdata = tidyr::expand_grid(
      log_dose = seq(from = lower, to = upper, length.out = 100),
      target = data$target %>% unique()),
    value = "response") %>%
  ggdist:: median_qi(.width = c(.5, .8, .95)) %>%
  dplyr::mutate(
    rmsd = exp(log_dose),
    score = response)

```

Visualize the fit as draws from the expected mean and median quantile intvervals on the
log(RMSD) scale:

```{r plot-data-model, dependson=c("load-packages", "stugdock-fit-model", "snugdock-collect-draws")}
ggplot2::ggplot(data = data) + 
  ggplot2::theme_bw() +
  ggplot2::facet_wrap(
    facets = dplyr::vars(target)) +
  ggplot2::scale_y_continuous("Interface Score (REU)") +
  ggplot2::scale_x_continuous(
    expression("Interface RMSD ("*ring(A)*")"),
    breaks = log(c(1, 2, 4, 8, 16)),
    labels = c(1, 2, 4, 8, 16)) +
  ggplot2::theme(legend.position = "bottom") +
  ggplot2::scale_fill_discrete(
    "Median Quantile Interval",
    labels = c("90%", "80%", "50%")) +
  ggplot2::ggtitle(
    "Snugdock Score vs. RMSD") +
  GeomIndicator::geom_indicator(
    data = pnear_scores,
    mapping = ggplot2::aes(
      indicator = Pnear_label),
    xpos = "left",
    group = 0,
    ypos = 0.80) +
  ggdist::geom_lineribbon(
    data = pp_data,
    mapping = ggplot2::aes(
      x = log(rmsd),
      y = score,
      ymin = .lower,
      ymax = .upper),
    alpha = 4/30) +
  ggplot2::geom_line(
    data = ep_data,
    mapping = ggplot2::aes(
      x = log(rmsd),
      y = score,
      group = .draw),
    color = "darkgreen",
    alpha = 20/30) +
  ggplot2::geom_point(
    mapping = ggplot2::aes(
      x = log(rmsd),
      y = score),
    size = .8,
    alpha = .8)

ggplot2::ggsave(
  filename = "two_state_fit.pdf",
  width = 7,
  height = 5,
  useDingbats = FALSE)


```
And on the original RMSD scale:


```{r plot-data-model-rmsd, dependson=c("load-packages", "stugdock-fit-model", "snugdock-collect-draws"), warning=FALSE}
ggplot2::ggplot(data = data) + 
  ggplot2::theme_bw() +
  GeomIndicator::geom_indicator(
    data = pnear_scores,
    mapping = ggplot2::aes(
      indicator = Pnear_label),
    xpos = "left",
    group = 0,
    ypos = 0.80) +
  ggdist::geom_lineribbon(
    data = pp_data,
    mapping = ggplot2::aes(
      x = rmsd,
      y = score,
      ymin = .lower,
      ymax = .upper),
    color = "darkgreen",
    alpha = 4/30) +
  ggplot2::geom_line(
    data = ep_data,
    mapping = ggplot2::aes(
      x = rmsd,
      y = score,
      group = .draw),
    color = "darkgreen",
    alpha = 20/30) +
  ggplot2::geom_point(
    mapping = ggplot2::aes(
      x = rmsd,
      y = score),
    alpha = 0.8,
    size = 0.8) +
  ggplot2::facet_wrap(
    facets = dplyr::vars(target)) +
  ggplot2::scale_y_continuous(
    "Interface Score (REU)") +
  ggplot2::scale_x_continuous(
    expression("Interface RMSD ("*ring(A)*")"),
    limits = c(1, 18.5),
    breaks = c(1, 4, 8, 12, 16)) +
  ggplot2::scale_fill_discrete(
    "Median Quantile Interval",
    labels = c("90%", "80%", "50%")) +
  ggplot2::theme(legend.position = "bottom") +
  ggplot2::ggtitle(
    "Snugdock Score vs. RMSD")

```


## Quantify Pnear uncertainty
Using the two-state model fit, it is possible to propagate the posterior uncertainty
to estimate the uncertainty of the Pnear.

```{r pnear-uncertainty}
# Median quantile intervals for the response across RMSD values for draws from the posterior
# pp_data <- model %>%
#   tidybayes::add_predicted_draws(
#     newdata = tidyr::expand_grid(
#       log_dose = seq(from = lower, to = upper, length.out = 100),
#       target = data$target %>% unique()),
#     value = "response") %>%
#   dplyr::group_by(target, .draw) %>%
#   dplyr::do({
#     draws <- .
#     data.frame(
#       
#     )
#   ggdist:: median_qi(.width = c(.5, .8, .95)) %>%
#   dplyr::mutate(
#     rmsd = exp(log_dose),
#     score = response)





