---
title: "BayesPharma: Bayesian methods for pharmacology models"
format:
  jss-pdf:
    keep-tex: true  
  jss-html: default
  docx:
    referenc-doc: template.docx
author:
  - name: Madeline J. Martin
    email: martin.mjm105@gmail.com
    affiliations:
      - ref: knight
  - name: Elayne Vieira Diaz
    email: Elayne.Vieiradias@ucsf.edu
    affiliations:
      - ref: weill
  - name: P. Walter German
    email: p.walter.german@gmail.com
    affiliations:
      - ref: weill
  - name: Elyssa B. Margolis
    email: elyssa.margolis@ucsf.edu
    orcid: 0000-0001-8777-302X
    affiliations:
      - ref: weill
      - ref: ucsf_ngp
  - name: "Matthew J. O'Meara"
    email: maom@umich.edu
    orcid: 0000-0002-3128-5331
    attributes:
      corresponding: true
    affiliations:
      - ref: umich_dcmb
      - ref: umich_medchem
affiliations:
  - id: knight
    name: University of Oregon
    department: Knight Campus Center for Accelerating Scientific Impact
    city: Eugene
    state: OR
    postal-code: 97403
    country: USA
  - id: weill
    name: University of California, San Francisco
    department: UCSF Weill Institute for Neurosciences, Department of Neurology 
    city: San Francisco
    state: CA
    postal-code: 94158
    country: USA
  - id: ucsf_ngp
    name: University of California, San Francisco
    department: Neuroscience Graduate Program
    city: San Francisco
    state: CA
    country: USA
  - id: umich_dcmb
    name: University of Michigan
    department: Department of Computational Medicine and Bioinformatics
    city: Ann Arbor
    state: MI
    country: USA
  - id: umich_medchem
    name: University of Michigan
    department: Department of Medichinal Chemistry
    city: Ann Arbor
    state: MI
    country: USA
highlight-style: github
abstract: |
  In pharmacology, many experiments seek to measure how a reductive biological
  system responds to one or more treatments. Here, we present `BayesPharma`, a
  collection of Bayesian methods to increase the robustness of the analysis and
  interpretation of dose response data. BayesPharma is an `R` package built
  around the `Stan` ecosystem that facilitates a principled Bayesian workflow.
  It can be used to fit and analyze foundational pharmacology models, serve as a
  pedagogical framework for learning Bayesian methods, and help build and
  analyze sophisticated pharmacological models. We describe the `BayesPharma`
  framework and illustrate its application through several case studies.
bibliography: references.bib
editor: visual
header-includes: |
  % https://colorbrewer2.org/#type=qualitative&scheme=Dark2&n=3
  \definecolor{dose}{rgb}{0.10546875,0.6171875,0.46484375}
  \definecolor{response}{rgb}{0.8476562,0.3710938,0.0078125}
  \definecolor{shape}{rgb}{0.4570312,0.4375000,0.6992188}
  \usepackage{chemarr}
knitr:
  opts_knit:
    verbose: true
  opts_chunk: 
    cache: true
    echo: false
    #fig.path: "../../vignettes_src/manuscript/manuscript_files/"
    fig.width: 6
    fig.height: 5
    fig.pos: "t"
    fig.align: "center"
    out.width: "100%"

keywords: [Bayesian, Pharmacology, R]
keywords-formatted: [Bayesian, Pharmacology, "[R]{.proglang}"]
---

```{r set-options}
#| echo=FALSE
# inspired by https://www.jumpingrivers.com/blog/knitr-default-options-settings-hooks/
knitr::opts_chunk$set(
  dpi = if (knitr::is_latex_output()) 72 else 300)
#  dev = "svg",
#  dev.args = list(png = list(type = "cairo-png")),
#  optipng = "-o1 -quiet")
# cmdstanr is more up-to-date than rstan and runs a little faster
# but can't expose defined functions for downstream analysis
if(Sys.info()["machine"] == "arm64"){
  #rstan does not currently support apple M1 chips
  stan_backend <- "cmdstanr"
} else {
  stan_backend <- "rstan"
}
```

```{r load-packages}
#suppressWarnings(suppressMessages(library(tidyverse)))
suppressWarnings(suppressMessages(library(BayesPharma)))
suppressWarnings(suppressMessages(library(drc)))
suppressWarnings(suppressMessages(library(brms)))
suppressWarnings(suppressMessages(library(broom)))
set.seed(0)
```

{{< include sections/introduction.qmd >}}

## Case Studies

In this section, we will consider several models as case studies: the sigmoidal
hill model in @sec-hill, the MuSyC synergy model in @sec-MuSyC}, and the
Michaelis-Menten enzyme progress curve in @sec-michaelis_menten}. For each, we
will implement it, apply it to example data by fitting different models, and
compare the models based on their fit of the data and inferences that can be
made.

{{< include sections/case_study_hill.qmd >}}

{{< include sections/case_study_synergy.qmd >}}

{{< include sections/case_study_kinetics.qmd >}}

{{< include sections/case_study_negbinom.qmd >}}

## Conclusion
The `BayesPharma` package aims to make Bayesian modeling for pharmacology more
accessible. To do this, it makes easy to use interfaces for foundational
dose-response type models. By building on the powerful Stan/BRMS framework,
BayesPharma can be light-weight and adaptable to more advanced use cases while
being computationally efficient and powerful. 

