---
title: "Apply: tQ Model -- ChABC Design"
description: Demonstrate the robust tQ model for enzyme kinetics of re-design
  of ChABC, an enzyme used to degrade extra cellular matrix.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Apply: tQ Model -- ChABC Design}
  %\VignetteEncoding{UTF-8}
---
  
```{r set-options}
#| echo=FALSE

# inspired by https://www.jumpingrivers.com/blog/knitr-default-options-settings-hooks/
knitr::opts_chunk$set(
  echo = FALSE,
  cache = TRUE,
  cache.path = "cache/apply_model_tQ_ChABC_design/",
  fig.path = "apply_model_tQ_ChABC_design_files/",
  fig.retina = 2, # Control using dpi
  fig.width = 6,  # generated images
  fig.height = 5, # generated images
  fig.pos = "t",  # pdf mode
  fig.align = "center",
  dpi = if (knitr::is_latex_output()) 72 else 300, 
  out.width = "100%")
```


```{r load-packages}
suppressWarnings(suppressMessages(library(tidyverse)))
library(BayesPharma)
library(GeomIndicator)
suppressWarnings(suppressMessages(library(tidybayes)))
```

Chondroitinase ABC (ChABC) is a bacterial enzyme that degrades several types of
proteoglycans. While natively, it is secreted to digest host tissue to
facilitate nutrient acquisition, it has been repurposed as a tool for glycomics
and as a therapeutic to treat scar formation, such post-stroke recovery. A
limitation of the widely used ChABC Proteus vulgaris for use in biotechnology
and medicine is that it is unstable and has a short half life. To overcome this,
several groups have tried to stabilize ChABC while maintaining or improving
enzyme efficiency. For example, in (Hettiaratchi, et al., 2020), they used
Protein One Stop Shop (PROSS) to generate 3 designs ranging from XX-YY number of
mutations and measured the enzyme efficiency of the enzyme to break down the
substrate Chondroitan A.



```{r read-data}
#| echo=FALSE
data_path <- here::here(
  "inst", "extdata", "enzyme_kinetics", "Hettiaratchi_2020.xlsx")

data <- tibble::tibble(
  sheet_name = readxl::excel_sheets(data_path)[2:7]) |>
  dplyr::rowwise() |>
  dplyr::do({
    sheet_name <- .$sheet_name
    data <- readxl::read_excel(
      path = data_path,
      sheet = sheet_name,
      range = "R2C2:R65C18",
      col_names = TRUE,
      .name_repair = ~ paste0("column_", 1:length(.))) |>
      dplyr::mutate(
        substrate_concentration_mg_ml = sheet_name |>
          stringr::str_extract("^[0-9.]+") |>
          as.numeric(),
          .before = 1)
  }) |>
  dplyr::ungroup()

names(data)[seq(4, 16, by = 3)] <- paste0(
  data[1,2:16] |> as.character() |> na.omit(), "_1")
names(data)[seq(5, 17, by = 3)] <- paste0(
  data[1,2:16] |> as.character() |> na.omit(), "_2")
names(data)[seq(6, 18, by = 3)] <- paste0(
  data[1,2:16] |> as.character() |> na.omit(), "_3")
names(data)[2] <- "time_s"
names(data)[3] <- "temperature_C"

data <- data |>
  dplyr::filter(
    !is.na(time_s),
    time_s != "Time [s]") |>
  tidyr::pivot_longer(
    cols = -c("substrate_concentration_mg_ml", "time_s", "temperature_C"),
    names_to = "treatment_replica",
    values_to = "product_concentration") |>
  tidyr::separate(
    col = treatment_replica,
    into = c("treatment", "replica"),
    sep = "_") |>
  dplyr::mutate(
    replica = as.factor(replica),
    time_s = as.numeric(time_s),
    temperature_C = as.numeric(temperature_C),
    product_concentration = ifelse(
      product_concentration == "OVER",
      NA, product_concentration),
    product_concentration = as.numeric(product_concentration))

baseline <- data |>
  dplyr::filter(treatment == "Blank") |>
  dplyr::group_by(substrate_concentration_mg_ml, time_s) |>
  dplyr::summarize(
    mean_baseline = mean(product_concentration, na.rm=TRUE),
    .groups = "drop")

data <- data |>
  dplyr::left_join(
    baseline,
    by = c("substrate_concentration_mg_ml", "time_s")) |>
  dplyr::mutate(
    normalized_product_concentration = product_concentration - mean_baseline)

```


```{r plot-data}
ggplot2::ggplot(
  data = data |>
    dplyr::filter(
      treatment != "Blank",
      !is.na(product_concentration))) + 
  ggplot2::theme_bw() +
  ggplot2::geom_line(
    mapping = ggplot2::aes(
      x = time_s,
      y = normalized_product_concentration,
      color = log(substrate_concentration_mg_ml),
      group = paste0(substrate_concentration_mg_ml, "_", replica))) +
  ggplot2::geom_smooth(
    data = data |>
      dplyr::filter(
        treatment != "Blank",
        time_s <= 200),
    mapping = ggplot2::aes(
      x = time_s,
      y = normalized_product_concentration,
      group = paste0(substrate_concentration_mg_ml, "_", replica)),
    method = "lm",
    formula = y ~ x,
    color = "orange") +
  ggplot2::facet_wrap(
    facets = dplyr::vars(treatment),
    nrow = 1) +
  ggplot2::theme(legend.position = "bottom") +
  ggplot2::scale_color_continuous(
    "Substrate mg/ml",
    breaks = log(c(0.5, 1, 2.5, 5, 10, 15)),
    labels = c("0.5", "1", "2.5", "5", "10", "15")) +
  ggplot2::scale_x_continuous("Time (s)") +
  ggplot2::scale_y_continuous(
    "Product mg/ml",
    breaks = c(0.5, 1, 2.5, 5, 10, 15),
    labels = c("0.5", "1", "2.5", "5", "10", "15")) 

```
```{r simulate-tQ}
data_simulate <- tidyr::expand_grid(
  kcat = 0.0072,
  kM =  0.0448,
  ET = 1,
  ST = c(0.5, 1, 2.5, 5, 10, 15)) |>
  dplyr::mutate(series_index = dplyr::row_number()) |>
  dplyr::rowwise() |>
  dplyr::do({
    data <- .
    time <- seq(0, 1200, by = 20)
    data <- data.frame(data,
      time = time,
      P = BayesPharma::tQ_model_generate(
        time = time,
        kcat = data$kcat,
        kM = data$kM,
        ET = data$ET,
        ST = data$ST)[,2])
  })

ggplot2::ggplot(data = data_simulate) + 
  ggplot2::theme_bw() +
  ggplot2::geom_line(
    mapping = ggplot2::aes(
      x = time,
      y = 0.8 + 0.16*P,
      color = log(ST),
      group = ST)) +
  ggplot2::theme(legend.position = "bottom") +
  ggplot2::scale_color_continuous(
    "Substrate mg/ml:",
    breaks = log(c(0.5, 1, 2.5, 5, 10, 15)),
    labels = c("0.5", "1", "2.5", "5", "10", "15")) +
  ggplot2::scale_x_continuous("Time (s)") +
  ggplot2::scale_y_continuous(
    "Product mg/ml:",
    breaks = c(0.5, 1, 2.5, 5, 10, 15),
    labels = c("0.5", "1", "2.5", "5", "10", "15"))



```

```{r model-tQ}
model_data <- data |>
  dplyr::arrange(
    treatment,
    substrate_concentration_mg_ml,
    replica,
    time_s)

model_data <- model_data |>
  dplyr::left_join(
    model_data |>
      dplyr::distinct(treatment, substrate_concentration_mg_ml, replica) |>
      dplyr::mutate(series_index = dplyr::row_number()),
    by = c("treatment", "substrate_concentration_mg_ml", "replica")) |>
  dplyr::filter(time_s > 0) |>
  dplyr::filter(!is.na(normalized_product_concentration)) |>
  dplyr::transmute(
    series_index,
    treatment,
    ET = 1,
    ST = substrate_concentration_mg_ml,
    P = product_concentration,
    time = time_s)

model_data <- model_data |>
  dplyr::filter(treatment != "Blank")

model_aa <- BayesPharma::tQ_model(
  data = model_data,
  formula = BayesPharma::tQ_formula(predictors = 0 + treatment),
  prior = BayesPharma::tQ_prior(
    kcat = brms::prior(prior = gamma(1, 200), lb = 0, nlpar = "kcat"),
    kM = brms::prior(prior = gamma(2, 0.2), lb = 0, nlpar = "kM")),
  init = BayesPharma::tQ_init(
    kcat = \() runif(n = 1, 0.001, 0.1),
    kM = \() runif(n = 1, 1, 20),
    sigma = \() runif(n = 1, 0, 2)),
  cores = 8,
  threads = 2,
  control = NULL,
  backend="cmdstanr",
  expose_functions = FALSE)



# trainable offset paramter
model_formula <- brms::brmsformula(
    P ~ tQ_multiple(series_index, time + offset, kcat, kM, ET, ST),
    kcat + kM ~ 0 + treatment,
    offset ~ 1,
    nl = TRUE,
    loop = FALSE)
model_formula$bayes_pharma_info <- list(
    formula_type = "tQ",
    series_index_variable = "series_index",
    treatment_variable = "time",
    treatment_units = "seconds",
    ET_variable = "ET",
    ET_units = "mg/ml",
    ST_variable = "ST",
    ST_units = "mg/ml",
    response_variable = "P",
    response_units = "mg/ml")


model_offset <- BayesPharma::tQ_model(
  data = model_data,
  formula = model_formula,
  prior = BayesPharma::tQ_prior(
    kcat = brms::prior(prior = gamma(1, 200), lb = 0, nlpar = "kcat"),
    kM = brms::prior(prior = gamma(2, 0.2), lb = 0, nlpar = "kM"),
    offset = brms::prior(prior = normal(0, 5), nlpar = "offset"),
    sigma = brms::prior(prior = student_t(3, 0, 2.5), class = "sigma")),
  init = list(
    b_kcat = \() runif(n = 1, 0.001, 0.1),
    b_kM = \() runif(n = 1, 1, 20),
    b_offset = \() runif(n = 1, -2, 2),
    sigma = \() runif(n = 1, 0, 2)) |>
    "class<-"("bpinit"),
  cores = 8,
  threads = 2,
  control = NULL,
  backend="cmdstanr",
  expose_functions = FALSE)

model_offset_rstan <- BayesPharma::tQ_model(
  data = model_data,
  formula = model_formula,
  prior = BayesPharma::tQ_prior(
    kcat = brms::prior(prior = gamma(1, 200), lb = 0, nlpar = "kcat"),
    kM = brms::prior(prior = gamma(2, 0.2), lb = 0, nlpar = "kM"),
    offset = brms::prior(prior = normal(0, 5), nlpar = "offset"),
    sigma = brms::prior(prior = student_t(3, 0, 2.5), class = "sigma")),
  init = list(
    b_kcat = \() runif(n = 1, 0.001, 0.1),
    b_kM = \() runif(n = 1, 1, 20),
    b_offset = \() runif(n = 1, -2, 2),
    sigma = \() runif(n = 1, 0, 2)) |>
    "class<-"("bpinit"),
  chains = 0,
  cores = 8,
  threads = 2,
  control = NULL,
  expose_functions = FALSE)


```

```{r plot-prior-posterior}

model_offset |>
  BayesPharma::plot_prior_posterior_densities()
```

```{r model-multiple-draws, deps=c("model-multiple", "model-multiple-prior")}

```

```{r prior-posterior-multiple-scatter, deps=c("model-multiple-draws")}
model <- model_offset

model_prior <- model |>
  update(sample_prior = "only")

draws_prior <- model_prior |>
  tidybayes::tidy_draws() |>
  tidybayes::gather_variables() |>
  dplyr::filter(.variable |> stringr::str_detect("b_(kcat|kM)")) |>
  tidyr::separate(.variable, into = c("class", "parameter", "treatment"), sep = "_") |>
  dplyr::mutate(treatment = treatment |> stringr::str_replace("treatment", ""))

draws_posterior <- model |>
  tidybayes::tidy_draws() |>
  tidybayes::gather_variables() |>
  dplyr::filter(.variable |> stringr::str_detect("b_(kcat|kM)")) |>
  tidyr::separate(.variable, into = c("class", "parameter", "treatment"), sep = "_") |>
  dplyr::mutate(treatment = treatment |> stringr::str_replace("treatment", ""))


draws_prior_pairs <- draws_prior |>
  tidyr::pivot_wider(
    id_cols = c(".chain", ".iteration", ".draw", "treatment"),
    names_from = "parameter",
    values_from = ".value")

draws_posterior_pairs <- draws_posterior |>
  tidyr::pivot_wider(
    id_cols = c(".chain", ".iteration", ".draw", "treatment"),
    names_from = "parameter",
    values_from = ".value")

ggplot2::ggplot() +
  ggplot2::theme_bw() +
    ggplot2::geom_point(
      data = draws_prior_pairs,
      mapping = ggplot2::aes(
        x = log10(kcat),
        y = log10(kM)),
      color = "orange",
      size = .8,
      shape = 16,
      alpha = .3) +  
    ggplot2::geom_point(
      data = draws_posterior_pairs,
      mapping = ggplot2::aes(
        x = log10(kcat),
        y = log10(kM)),
      color = "blue",
      size = .8,
      shape = 16,
      alpha = .3) +
  ggplot2::scale_x_continuous("Catalytic constant log10(kcat)") +
  ggplot2::scale_y_continuous("Michaelis constant log10(kM)") +
  ggplot2::facet_wrap(facets = dplyr::vars(treatment), nrow = 1)
```



```{r posterior-draws}
model <- model_offset


newdata_levels <- model$data |>
    dplyr::select(tidyselect::any_of(c("treatment", "ET", "ST", "time"))) |>
    as.list() |>
    purrr::map(unique)
newdata_levels$time = seq(-120, 1200, 100)
newdata <- do.call(tidyr::expand_grid, newdata_levels)

# add series_index
newdata <- newdata |>
  dplyr::left_join(
    newdata |>
      dplyr::distinct(treatment, ET, ST) |>
      dplyr::mutate(series_index = dplyr::row_number()),
    by = c("treatment", "ET", "ST"))

pp_data <- model |>
  tidybayes::add_predicted_draws(
    newdata = newdata,
    value = "P",
    re_formula = NA,
    ndraws = 200)  |>
  ggdist:: median_qi(.width = c(.5, .8, .95))

ggplot2::ggplot() +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom") +
  ggplot2::scale_fill_discrete(
    "Median Quantile Interval",
    labels = c("95%", "80%", "50%")) +
  ggdist::geom_lineribbon(
    data = pp_data,
    mapping = ggplot2::aes(
      x = .data[["time"]],
      y = .data[["P"]],
      ymin = .data[[".lower"]],
      ymax = .data[[".upper"]],
      color = log10(.data[["ST"]]),
      group = .data[["ST"]]),
    alpha = .15) +
  ggplot2::geom_point(
    data = model$data,
    mapping = ggplot2::aes(
      x = .data[["time"]],
      y = .data[["P"]],
      color = log10(.data[["ST"]])),
    size = 0.8,
    alpha = 0.7) +
  ggplot2::facet_wrap(facets = ~treatment) +
  ggplot2::labs(title = "Predicted tQ Enzyme Progress Curve") +
  ggplot2::scale_color_viridis_c(
    "Substrate Concentration (uM)",
    labels = c("0.5", "2.5", "15"),
    breaks = log10(c(0.5, 2.5, 15)),
    begin = 0.2, end = 0.9) +
  ggplot2::scale_y_continuous("Product Concentration") +
  ggplot2::scale_x_continuous("Time (s)")

ggplot2::ggsave(
  filename = "/tmp/model_offset_posterior_draws.pdf",
  width = 10,
  height = 6,
  useDingbats = FALSE)

```
