---
# For vignette -----------------------------------------------------------------
title: "Derive: MuSyC Model -- Synergy Analysis"
description: Derive the functional form, parameterization, and implementation
  for the MuSyC synergy model.
# Document ---------------------------------------------------------------------
output: rmarkdown::html_vignette
bibliography: references.bib
editor: visual
header-includes: |
  \usepackage{chemarr}
vignette: >
  %\VignetteIndexEntry{Derive: MuSyC Model -- Synergy Analysis}
  %\VignetteKeyword{BayesPharma}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r set-options}
#| echo=FALSE
options(rmarkdown.html_vignette.check_title = FALSE)

# inspired by https://www.jumpingrivers.com/blog/knitr-default-options-settings-hooks/
knitr::opts_chunk$set(
  echo = FALSE,
  cache = TRUE,
  cache.path = "cache/derive_MuSyC_model/",
  fig.path = "derive_MuSyC_model_files/",
  fig.retina = 2, # Control using dpi
  fig.width = 6,  # generated images
  fig.height = 5, # generated images
  fig.pos = "t",  # pdf mode
  fig.align = "center",
  dpi = if (knitr::is_latex_output()) 72 else 300, 
  out.width = "100%",
  collapse = TRUE)

set.seed(0)
```

```{r load-packages}
#| echo=FALSE
suppressWarnings(suppressMessages(library(tidyverse)))
library(BayesPharma)

bayesplot::bayesplot_theme_set(
  new = ggplot2::theme_bw())
```

```{r}
#| child = here::here("vignettes_src", "manuscript", "sections", "case_study_synergy.qmd")

```

\#' Drug Synergy \#' MuSyC Drug Synergy model \#' \#' Assume that the response metric decreases with more effective drugs \#' Let E3 be the effect at the maximum concentration of both drugs \#' \#' \#' Special cases: \#' \* dose additive model: alpha1 = alpha2 = 0 \#' \* loewe: h1 = h2 = 1 \#' \* CI: E0 = 1, E1 = E2 = E3 = 0 \#' the drug effect is equated with percent inhibition \#' \* bliss drug independence model: \#' E0 = 1, E1 = E2 = E3 = 0, alpha1 = alpha2 = 1 \#' @param d1 Dose of drug 1 \#' @param d2 Dose of drug 2 \#' \#' @param E0 effect with no drug treatment \#' \#' \# params for drug 1 by it self \#' @param s1 drug 1 hill slope \#' @param C1 drug 1 EC50 \#' @param E1 drug 1 maximum effect \#' \#' \# params for drug 2 by it self \#' @param s2 drug 2 hill slope \#' @param C2 drug 2 EC50 \#' @param E2 drug 2 maximum effect \#' \#' @param beta synergistic efficacy \#' percent increase in a drug combination's effect \#' beyond the most efficacious single drug. \#' \#' beta \> 0 =\> synergistic efficacy \#' the effect at the maximum concentration of both drugs (E3) exceeds the \#' maximum effect of either drug alone (E1 or E2) \#' \#' beta \< 0 =\> antagonistic efficacy \#' at least one or both drugs are more efficacious as \#' single agents than in combination \#' \#' @param alpha1 synergistic potency \#' how the effective dose of drug 1 \#' is altered by the presence of drug 2 \#' @param alpha2 synergistic potency \#' how the effective dose of drug 2 \#' is altered by the presence of drug 1 \#' \#' alpha \> 1 =\> synergistic potency \#' the EC50 decreases because of the addition of the other drug, \#' corresponding to an increase in potency \#' \#' 0 \<= alpha \< 1 =\> antagonistic potency \#' the EC50 of the drug increases as a result of the other drug, \#' corresponding to a decrease in potency \#' \#' alpha1 == alpha2 if detailed balance \#' @export generate_MuSyC_effects \<- function( d1, d2, E0, s1, C1, E1, s2, C2, E2, alpha, E3) { h1 \<- MuSyC_si_to_hi(s1, C1, E0, E1) h2 \<- MuSyC_si_to_hi(s2, C2, E0, E2) numerator \<- C1\^h1 \* C2\^h2 \* E0 + d1\^h1 \* C2\^h2 \* E1 + C1\^h1 \* d2\^h2 \* E2 + d1\^h1 \* d2\^h2 \* E3 \* alpha denominator \<- C1\^h1 \* C2\^h2 + d1\^h1 \* C2\^h2 + C1\^h1 \* d2\^h2 + d1\^h1 \* d2\^h2 \* alpha numerator / denominator }

\#' Create a formula for the MuSyC synergy model \#' \#' @description setup a defaulMuSyC synergy model formula to predict \#' the `E0`, `C1`, `E1`, `s1`, `C2`, `E2`, `s2`, `log10alpha`, and `E3alpha` \#' parameters. \#' \#' @param predictors Additional formula objects to specify predictors of \#' non-linear parameters. i.e. what perturbations/experimental differences \#' should be modeled separately? (Default: 1) should a random effect be taken \#' into consideration? i.e. cell number, plate number, etc. \#' @return brmsformula \#' \#' @examples \#'\dontrun{
#'   Consider observations made using 4 different drugs and the column header
#'   containing the labels for the 4 different drugs is `predictors`.
#'   
#'      `MuSyC_formula(predictors = 0 + predictors)`
#'
#'   Consider that the cell_ID was recorded and the noise from the cell_ID is to
#'   be accounted for.
#'  
#'      `MuSyC_formula(predictors = 0 + predictors + (1|cell_ID))`
#'} \#' \#' @export MuSyC_formula \<- function( predictors = 1, ...) {

```         
predictor_eq <- rlang::new_formula(
  lhs = quote(E0 + C1 + E1 + s1 + C2 + E2 + s2 + log10alpha + E3alpha),
  rhs = rlang::enexpr(predictors))

brms::brmsformula(
  response ~ (C1^h1 * C2^h2 * E0 +
      d1^h1 * C2^h2 * E1 +
      C1^h1 * d2^h2 * E2 +
      d1^h1 * d2^h2 * E3alpha
    ) / (
      C1^h1 * C2^h2 +
      d1^h1 * C2^h2 +
      C1^h1 * d2^h2 +
      d1^h1 * d2^h2 * 10^log10alpha),
  brms::nlf(d1 ~ dose1 / d1_scale_factor),
  brms::nlf(d2 ~ dose2 / d2_scale_factor),
  brms::nlf(h1 ~ s1 * (4 * C1) / (E0 + E1)),
  brms::nlf(h2 ~ s2 * (4 * C2) / (E0 + E2)),
  predictors_eq,
  nl = TRUE,
  ...)
```

}

\#' Fit the MuSyC synergy model by dose \#' \#' @param data data.frame of experimental data \#' with columns: dose1, dose2, n_positive, count, [<group_vars>] \#' @param group_vars quosures list \#' dplyr::vars(...) columns to over when fitting synergy model \#' @param C1_prior prior distribution for Ed when d1=d1_IC50, d2=0 \#' @param C2_prior prior distribution for Ed when d1=0, d2=d2_IC50 \#' @param s1_prior prior distribution for d(Ed)/d(d1) when d1=d1_IC50, d2=0 \#' @param s2_prior prior distribution for d(Ed)/d(d2) when d1=0, d2=d2_IC50 \#' @param log10alpha_prior prior distribution for alpha synergy parameter \#' @param E0_prior prior distribution for Ed when d1=0, d2=0 \#' @param E1_prior prior distribution for Ed when d1=Inf, d2=0 \#' @param E2_prior prior distribution for Ed when d1=0, d2=Inf \#' @param E3_alpha_prior prior distribution for Ed scaled by alpha when d1=Inf, \#' d2=Inf \#' @param C1_init initial sampling distribution for the C1 parameter \#' @param C2_init initial sampling distribution for the C2 parameter \#' @param s1_init initial sampling distribution for the s1 parameter \#' @param s2_init initial sampling distribution for the s2 parameter \#' @param log10alpha_init initial sampling distribution for the alpha parameter \#' @param E0_init initial sampling distribution for the E0 parameter \#' @param E1_init initial sampling distribution for the E1 parameter \#' @param E2_init initial sampling distribution for the E2 parameter \#' @param E3_alpha_init initial sampling distribution for the E3 parameter \#' @param combine combine the grouped models into a single brms model \#' @param verbose verbose output \#' \#' @param iter number of stan NUTS sampling steps \#' @param cores number of cores used for sampling \#' @param stan_model_args stan model arguments \#' @param control stan control arguments \#' \#' The \#' \#' bernoulli_inf(n_positive / count) = \#' Ed \~ MuSyC(d1, d2, C_params, E_params, s_params, alpha) \#' \#' To improve numeric stability, the d1 and d2 and C1 and C2 variables \#' are scaled to improve numeric stability: \#' \#' d1 = dose1/max(dose1) \#' d2 = dose2/max(dose2) \#' drug1_IC50 = C1 \* max(dose1) \#' drug2_IC50 = C2 \* max(dose2) \#' \#' Functional form: \#' Ed \~ ( \#' C1\^h1 \* C2\^h2 \* E0 + \#' d1\^h1 \* C2\^h2 \* E1 + \#' C1\^h1 \* d2\^h2 \* E2 + \#' d1\^h1 \* d2\^h2 \* E3 \* alpha \#' ) / ( \#' C1\^h1 \* C2\^h2 + \#' d1\^h1 \* C2\^h2 + \#' C1\^h1 \* d2\^h2 + \#' d1\^h1 \* d2\^h2 \* alpha \#' ) \#' \#' \#' \#' \#' \#' \############################################## \#' \# Proof of the definitions of the parameters \# \#' \############################################## \#' \#' Claim: When d1=0 and d2=0 then Ed = E0 \#' Ed = (C1\^h1 \* C2\^h2 \* E0) / (C1\^h1 \* C2\^h2) \#' = E0 \#' \#' Claim: When d1=0 and d2 -\> Inf then Ed = E2 \#' Ed = (C2\^h2 \* E0 + d2\^h2 \* E2) / (C2\^h2 + d2\^h2) \#' = (d2\^h2 \* E2) / (d2\^h2) \#' = E2 \#' #; Claim: When d1=0 and d2=C2 then Ed = (E0 + E2) / 2 \#' When d1\>0 and d2 -\> Inf then Ed \#' Ed = (C1\^h1 \* C2\^h2 \* E0 + C1\^h1 \* C2\^h2 \* E2) / \#' (C1\^h1 \* C2\^h2 + C1\^h1 \* C2\^h2) \#' = (E0 + E2) / 2 \#' \#'@export MuSyC_model \<- function( data, group_vars = vars(compound), formula = MuSyC_formula(), prior = MuSyC_prior(), init = MuSyC_init(), combine = FALSE, verbose = FALSE, iter = 8000, cores = 4, stan_model_args = list(verbose = FALSE), control = list( adapt_delta = .99, max_treedepth = 12), model_evaluation_criteria = c("loo", "bayes_R2"), ...) {

if (is.data.frame(well_scores)) { grouped_data \<- well_scores \|\> dplyr::group_by(!!!group_vars) \|\> dplyr::mutate( d1_scale_factor = max(dose1), d2_scale_factor = max(dose2)) \|\> tidyr::nest() \|\> dplyr::ungroup() }

if (verbose) { cat("Fitting MuSyC model\n") }

model \<- brms::brm_multiple( formula = formula, data = grouped_data\$data, family = binomial("identity"), prior = prior, init = init, \# stanvars = c( \# brms::stanvar( \# scode = " real d1_scale_factor = max(dose1));", \# block ="tdata", \# position = "end"), \# brms::stanvar( \# scode = " real d2_scale_factor = max(dose2));", \# block ="tdata", \# position = "end"), \# brms::stanvar( \# scode = " real drug1_IC50 = b_C1 \* d1_scale_factor);", \# block ="genquant", \# position = "end"), \# brms::stanvar( \# scode = " real drug2_IC50 = b_C2 \* d2_scale_factor;", \# block ="genquant", \# position = "end")), combine = FALSE, data2 = NULL, iter = iter, cores = cores, stan_model_args = stan_model_args, control = control, ...)

if (!is.null(model_evaluation_criteria)) { \# evaluate fits model \<- model \|\> purrr::imap(function(model, i) { group_index \<- grouped_data[i, ] \|\> dplyr::select(-data) group_index_label \<- paste0( names(group_index), ":", group_index, collapse = ",") cat("Evaluating model fit for", group_index_label, "...\n", sep = "") model \<- model \|\> brms::add_criterion( criterion = model_evaluation_criteria, model_name = paste0("MuSyC:", group_index_label), reloo = TRUE) model }) } grouped_data \|\> dplyr::mutate( model = model) }
